---
format: gfm
default-image-extension: ""
editor_options: 
  chunk_output_type: console
---

# mlsurvlrnrs

<!-- badges: start -->
```{r}
#| echo: false
#| message: false
#| results: asis

pkg <- desc::desc_get_field("Package")
cat_var <- paste(
  badger::badge_lifecycle(),
  badger::badge_cran_release(pkg = pkg),
  gsub("summary", "worst", badger::badge_cran_checks(pkg = pkg)),
  badger::badge_cran_download(pkg = pkg, type = "grand-total", color = "blue"),
  badger::badge_cran_download(pkg = pkg, type = "last-month", color = "blue"),
  gsub("netlify\\.com", "netlify.app", badger::badge_dependencies(pkg = pkg)),
  badger::badge_github_actions(action = utils::URLencode("R CMD Check via {tic}")),
  badger::badge_github_actions(action = "lint"),
  badger::badge_github_actions(action = "test-coverage"),
  badger::badge_codecov(ref = desc::desc_get_urls()),
  sep = "\n"
)
cat_var |> cat()
```
<!-- badges: end -->

The goal of `mlsurvlrnrs` is to enhance the [`mlexperiments`](https://github.com/kapsner/mlexperiments) R package with survival learners. 

Currently implemented learners are:

| Name | Based on | Description / Tasks |
| ---- | -------- | ------------------- |
| LearnerSurvCoxPHCox | `survival::coxph` | Cox Proportional Hazards Regression |
| LearnerSurvGlmnetCox | `glmnet::glmnet` | Regularized Cox Regression |
| LearnerSurvRangerCox | `ranger::ranger` | Random Survival Forest with right-censored data |
| LearnerSurvRpartCox | `rpart::rpart` | Random Survival Forest with right-censored data |
| LearnerSurvXgboostCox | `xgboost::xgb.train` | Cox Regression with right-censored data |
| LearnerSurvXgboostAft | `xgboost::xgb.train` | [Accelerated failure time models](https://xgboost.readthedocs.io/en/stable/tutorials/aft_survival_analysis.html) with right-censored data |

For a short introduction on how to use the learners together with the `mlexperiments` R package, please visit the [wiki page](https://github.com/kapsner/mlsurvlrnrs/wiki).

## Installation

To install the development version, run

```{r}
#| eval: false
install.packages("remotes")
remotes::install_github("kapsner/mlsurvlrnrs")
```
