---
title: "survivalsvm: Survival Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{survivalsvm: Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
# nolint start
```

```{r setup}
library(mlexperiments)
library(mlsurvlrnrs)
```

See [https://github.com/kapsner/mlsurvlrnrs/blob/main/R/learner_surv_survivalsvm.R](https://github.com/kapsner/mlsurvlrnrs/blob/main/R/learner_surv_survivalsvm.R) for implementation details.

# Preprocessing

## Import and Prepare Data

```{r}
dataset <- survival::colon |>
  data.table::as.data.table() |>
  na.omit()
dataset <- dataset[get("etype") == 2, ]

surv_cols <- c("status", "time", "rx")
feature_cols <- colnames(dataset)[3:(ncol(dataset) - 1)]
```

## General Configurations

```{r}
seed <- 123
if (isTRUE(as.logical(Sys.getenv("_R_CHECK_LIMIT_CORES_")))) {
  # on cran
  ncores <- 2L
} else {
  ncores <- ifelse(
    test = parallel::detectCores() > 4,
    yes = 4L,
    no = ifelse(
      test = parallel::detectCores() < 2L,
      yes = 1L,
      no = parallel::detectCores()
    )
  )
}
options("mlexperiments.bayesian.max_init" = 10L)
```

## Generate Training- and Test Data

```{r}
split_vector <- splitTools::multi_strata(
  df = dataset[, .SD, .SDcols = surv_cols],
  strategy = "kmeans",
  k = 4
)

data_split <- splitTools::partition(
  y = split_vector,
  p = c(train = 0.7, test = 0.3),
  type = "stratified",
  seed = seed
)

train_x <- model.matrix(
  ~ -1 + .,
  dataset[
    data_split$train, .SD, .SDcols = setdiff(feature_cols, surv_cols[1:2])
  ]
)
train_y <- survival::Surv(
  event = (dataset[data_split$train, get("status")] |>
             as.character() |>
             as.integer()),
  time = dataset[data_split$train, get("time")],
  type = "right"
)
split_vector_train <- splitTools::multi_strata(
  df = dataset[data_split$train, .SD, .SDcols = surv_cols],
  strategy = "kmeans",
  k = 4
)


test_x <- model.matrix(
  ~ -1 + .,
  dataset[data_split$test, .SD, .SDcols = setdiff(feature_cols, surv_cols[1:2])]
)
test_y <- survival::Surv(
  event = (dataset[data_split$test, get("status")] |>
             as.character() |>
             as.integer()),
  time = dataset[data_split$test, get("time")],
  type = "right"
)
```

## Generate Training Data Folds

```{r}
fold_list <- splitTools::create_folds(
  y = split_vector_train,
  k = 3,
  type = "stratified",
  seed = seed
)
```

# Experiments

## Prepare Experiments

```{r}
# required learner arguments, not optimized
learner_args <- list(
  type = "regression",
  opt.meth = "quadprog",
  kernel = "rbf_kernel",
  maxiter = 5
)

# set arguments for predict function and performance metric,
# required for mlexperiments::MLCrossValidation and
# mlexperiments::MLNestedCV
predict_args <- NULL
performance_metric <- c_index
performance_metric_args <- NULL
return_models <- FALSE

# required for grid search and initialization of bayesian optimization
parameter_grid <- expand.grid(
  gamma.mu = seq(0.1, 0.9, 0.2)
)

# reduce to a maximum of 10 rows
if (nrow(parameter_grid) > 10) {
  set.seed(123)
  sample_rows <- sample(seq_len(nrow(parameter_grid)), 10, FALSE)
  parameter_grid <- kdry::mlh_subset(parameter_grid, sample_rows)
}

# required for bayesian optimization
parameter_bounds <- list(
  gamma.mu = c(0., 1.)
)
optim_args <- list(
  iters.n = ncores,
  kappa = 3.5,
  acq = "ucb"
)
```


## Hyperparameter Tuning

### Grid Search

```{r}
tuner <- mlexperiments::MLTuneParameters$new(
  learner = LearnerSurvSurvivalsvm$new(),
  strategy = "grid",
  ncores = ncores,
  seed = seed
)

tuner$parameter_grid <- parameter_grid
tuner$learner_args <- learner_args
tuner$split_type <- "stratified"
tuner$split_vector <- split_vector_train

tuner$set_data(
  x = train_x,
  y = train_y
)

tuner_results_grid <- tuner$execute(k = 3)
#> 
#> Parameter settings [=======================================================>------------------------------------------------------------------------------------] 2/5 ( 40%)
#> Parameter settings [===================================================================================>--------------------------------------------------------] 3/5 ( 60%)
#> Parameter settings [===============================================================================================================>----------------------------] 4/5 ( 80%)
#> Parameter settings [============================================================================================================================================] 5/5 (100%)                                                                                                                                                                             

head(tuner_results_grid)
#>    setting_id metric_optim_mean gamma.mu       type opt.meth     kernel maxiter
#> 1:          1         0.4226923      0.1 regression quadprog rbf_kernel       5
#> 2:          2         0.4226923      0.3 regression quadprog rbf_kernel       5
#> 3:          3         0.4226493      0.5 regression quadprog rbf_kernel       5
#> 4:          4         0.4226389      0.7 regression quadprog rbf_kernel       5
#> 5:          5         0.4226698      0.9 regression quadprog rbf_kernel       5
```


### Bayesian Optimization

```{r}
tuner <- mlexperiments::MLTuneParameters$new(
  learner = LearnerSurvSurvivalsvm$new(),
  strategy = "bayesian",
  ncores = ncores,
  seed = seed
)

tuner$parameter_grid <- parameter_grid
tuner$parameter_bounds <- parameter_bounds

tuner$learner_args <- learner_args
tuner$optim_args <- optim_args

tuner$split_type <- "stratified"
tuner$split_vector <- split_vector_train

tuner$set_data(
  x = train_x,
  y = train_y
)

tuner_results_bayesian <- tuner$execute(k = 3)
#> 
#> Registering parallel backend using 4 cores.

head(tuner_results_bayesian)
#>    Epoch setting_id gamma.mu gpUtility acqOptimum inBounds Elapsed     Score metric_optim_mean                            errorMessage       type opt.meth     kernel maxiter
#> 1:     0          1      0.1        NA      FALSE     TRUE  41.423 0.4226923         0.4226923                                    <NA> regression quadprog rbf_kernel       5
#> 2:     0          2      0.3        NA      FALSE     TRUE  41.794 0.4226923         0.4226923                                    <NA> regression quadprog rbf_kernel       5
#> 3:     0          3      0.5        NA      FALSE     TRUE  41.788 0.4226493         0.4226493                                    <NA> regression quadprog rbf_kernel       5
#> 4:     0          4      0.7        NA      FALSE     TRUE  41.020 0.4226389         0.4226389                                    <NA> regression quadprog rbf_kernel       5
#> 5:     0          5      0.9        NA      FALSE     TRUE  19.455 0.4226698         0.4226698                                    <NA> regression quadprog rbf_kernel       5
#> 6:     1          6      0.0 0.6397671       TRUE     TRUE   0.002        NA                NA gamma.mu: only positive values allowed. regression quadprog rbf_kernel       5
```

## k-Fold Cross Validation

```{r}
validator <- mlexperiments::MLCrossValidation$new(
  learner = LearnerSurvSurvivalsvm$new(),
  fold_list = fold_list,
  ncores = ncores,
  seed = seed
)

validator$learner_args <- tuner$results$best.setting[-1]

validator$predict_args <- predict_args
validator$performance_metric <- performance_metric
validator$performance_metric_args <- performance_metric_args
validator$return_models <- return_models

validator$set_data(
  x = train_x,
  y = train_y
)

validator_results <- validator$execute()
#> 
#> CV fold: Fold1
#> 
#> CV fold: Fold2
#> CV progress [=================================================================================================>-------------------------------------------------] 2/3 ( 67%)
#> 
#> CV fold: Fold3
#> CV progress [===================================================================================================================================================] 3/3 (100%)
#>                                                                                                                                                                              

head(validator_results)
#>     fold performance     gamma.mu       type opt.meth     kernel maxiter
#> 1: Fold1   0.4487941 0.0009792652 regression quadprog rbf_kernel       5
#> 2: Fold2   0.4036467 0.0009792652 regression quadprog rbf_kernel       5
#> 3: Fold3   0.4529986 0.0009792652 regression quadprog rbf_kernel       5
```


## Nested Cross Validation

### Inner Grid Search

```{r}
validator <- mlexperiments::MLNestedCV$new(
  learner = LearnerSurvSurvivalsvm$new(),
  strategy = "grid",
  fold_list = fold_list,
  k_tuning = 3L,
  ncores = ncores,
  seed = seed
)

validator$parameter_grid <- parameter_grid
validator$learner_args <- learner_args
validator$split_type <- "stratified"
validator$split_vector <- split_vector_train

validator$predict_args <- predict_args
validator$performance_metric <- performance_metric
validator$performance_metric_args <- performance_metric_args
validator$return_models <- return_models

validator$set_data(
  x = train_x,
  y = train_y
)

validator_results <- validator$execute()
#> 
#> CV fold: Fold1
#> 
#> Parameter settings [=======================================================>------------------------------------------------------------------------------------] 2/5 ( 40%)
#> Parameter settings [===================================================================================>--------------------------------------------------------] 3/5 ( 60%)
#> Parameter settings [===============================================================================================================>----------------------------] 4/5 ( 80%)
#> Parameter settings [============================================================================================================================================] 5/5 (100%)                                                                                                                                                                             
#> CV fold: Fold2
#> CV progress [=================================================================================================>-------------------------------------------------] 2/3 ( 67%)
#> 
#> Parameter settings [=======================================================>------------------------------------------------------------------------------------] 2/5 ( 40%)
#> Parameter settings [===================================================================================>--------------------------------------------------------] 3/5 ( 60%)
#> Parameter settings [===============================================================================================================>----------------------------] 4/5 ( 80%)
#> Parameter settings [============================================================================================================================================] 5/5 (100%)                                                                                                                                                                             
#> CV fold: Fold3
#> CV progress [===================================================================================================================================================] 3/3 (100%)
#>                                                                                                                                                                              
#> Parameter settings [=======================================================>------------------------------------------------------------------------------------] 2/5 ( 40%)
#> Parameter settings [===================================================================================>--------------------------------------------------------] 3/5 ( 60%)
#> Parameter settings [===============================================================================================================>----------------------------] 4/5 ( 80%)
#> Parameter settings [============================================================================================================================================] 5/5 (100%)                                                                                                                                                                             

head(validator_results)
#>     fold performance gamma.mu       type opt.meth     kernel maxiter
#> 1: Fold1   0.4365956      0.1 regression quadprog rbf_kernel       5
#> 2: Fold2   0.3917775      0.1 regression quadprog rbf_kernel       5
#> 3: Fold3   0.4397037      0.1 regression quadprog rbf_kernel       5
```


### Inner Bayesian Optimization

```{r}
validator <- mlexperiments::MLNestedCV$new(
  learner = LearnerSurvSurvivalsvm$new(),
  strategy = "bayesian",
  fold_list = fold_list,
  k_tuning = 3L,
  ncores = ncores,
  seed = 312
)

validator$parameter_grid <- parameter_grid
validator$learner_args <- learner_args
validator$split_type <- "stratified"
validator$split_vector <- split_vector_train


validator$parameter_bounds <- parameter_bounds
validator$optim_args <- optim_args

validator$predict_args <- predict_args
validator$performance_metric <- performance_metric
validator$performance_metric_args <- performance_metric_args
validator$return_models <- TRUE

validator$set_data(
  x = train_x,
  y = train_y
)

validator_results <- validator$execute()
#> 
#> CV fold: Fold1
#> 
#> Registering parallel backend using 4 cores.
#> 
#> CV fold: Fold2
#> CV progress [=================================================================================================>-------------------------------------------------] 2/3 ( 67%)
#> 
#> Registering parallel backend using 4 cores.
#> 
#> CV fold: Fold3
#> CV progress [===================================================================================================================================================] 3/3 (100%)
#>                                                                                                                                                                              
#> Registering parallel backend using 4 cores.

head(validator_results)
#>     fold performance    gamma.mu       type opt.meth     kernel maxiter
#> 1: Fold1   0.4364907 0.909125538 regression quadprog rbf_kernel       5
#> 2: Fold2   0.3917457 0.036876500 regression quadprog rbf_kernel       5
#> 3: Fold3   0.4530633 0.001528976 regression quadprog rbf_kernel       5
```


## Holdout Test Dataset Performance

### Predict Outcome in Holdout Test Dataset

```{r}
preds_survivalsvm <- mlexperiments::predictions(
  object = validator,
  newdata = test_x
)
```

### Evaluate Performance on Holdout Test Dataset

```{r}
perf_survivalsvm <- mlexperiments::performance(
  object = validator,
  prediction_results = preds_survivalsvm,
  y_ground_truth = test_y
)
perf_survivalsvm
#>    model performance
#> 1: Fold1   0.4402035
#> 2: Fold2   0.4582914
#> 3: Fold3   0.4755862
```

```{r include=FALSE}
# nolint end
```
